Predicting Remaining Useful Life of NASA Aircraft Engines

### 2025-03-08 index.qmd
updated kaplan meier section

To determine whether to use CPH for modeling, the data must contain a
time to event value and meet various assumptions, including proportional
hazards. Data that is censored will require a survival analysis that
type of analysis is deliberately designed for censored data. The data
must also have a binary indicator variable for the event occurrence and
no missing values [@nahhas2025].

@@

This equation can then be adapted for the specific dataset being
studied.

@@

**Kaplan-Meier**

Before generating the CPH model, it is beneficial to look at the
Kaplan-Meier curve. The Kaplan-Meier estimate ...

@@

To generate the CPH model, a survival object must first be created.
Using the R package `survival`, the suvival formula can be generated by
running `Surv(time, event) ~ x`. That formula is then applied to the R
function `coxph(formula, data=)` which will generated the CPH model
[@R-base].

@@

For the CPH model to be valid, the following assumptions must be met. If
they are not met, the data must either be transformed.

@@

@@

*Notes:*

*How coxph was created and creator what it was made for. it is semi
parametric regression*

*explain mathematically with equation*

*How data needs to be formatted - time + event and covariates*

*how it works*

*assumptions - how to test assumptions*

-   *Detail the models or algorithms used.*

-   *Justify your choices based on the problem and data.*

@@

```{r, echo = FALSE}
#packages
library(tidyverse)
library(gridExtra)
library(survival)
library(survminer)
library(ggpubr)
library(magrittr)
library(ggfortify)
library(knitr)

@@

Example for one predictor
```{r}
martingale <- resid(cox, type = "martingale")

>>>>

### Data Requirements

CPH requires data to contain a time to event value and have a binary
indicator variable for the event occurrence. The data must not contain
any missing values, but can be data that is censored which CPH is
designed to handle [@nahhas2025]. In addition, the data must meet
several assumptions for CPH to be a valid model to use on the data which
are discussed below.

@@

This equation can then be adapted for the specific data being studied.

@@

**Kaplan-Meier Estimator and Curve**

The Kaplan-Meier estimator and curve are frequently used in survival
analysis, especially when dealing with censored data or data where the
time-to-event value fluctuates per subject [@klein2005]. The
Kaplan-Meier estimate was proposed by Edward L. Kaplan and Paul Meier in
1958 to deal with incomplete observations [@rich2010]. It is the
standard estimator of the survival function and is defined by the
following equation:

$$
\hat{S}(t) = \begin{cases}
1 & \text{if }\; t < t_1, \\
\prod_{t_i \leq t}[1-\frac{d_i}{Y_i}], & \text{if }\; t_1 \leq t
\end{cases}
$$

-   $d_i$ is the number of events at time $t_i$
-   $Y_i$ is the number of individuals who are at risk at time $t_i$

for all values of $t$ where there is data. The estimator is not
well-defined for $t$ beyond the largest data value for time
[@klein2005].

The Kaplan-Meier curve on the X-axis represents the length of survival
time on the interval from 0 to the largest observation time while the
Y-axis represents the cumulative probability of surviving at a given
time. The curve is based on step-wise estimates thus making the curve
non-continuous [@rich2010]. Generating a Kaplan-Meier curve requires the
R packages `survival` and `survminer` and uses the survival object and
survival formula for the data. The below curve was generated using the
`survival` package and dataset "lung" and demonstrates the general
survival probability.

```{r}
library(survival)
library(survminer)
ggsurvplot(survfit(Surv(time, status) ~ 1, data = lung))

Kaplan-Meier can be used to compare the survival probability of
different groups, such as the sex of an individual. The log rank test
can be used to determine the significance of each group. The test
calculates the $\chi^2$ value for each event time for each group and
sums the results which are then added together to generate the $\chi^2$
value for the full curve for each group [@rich2010]. The hypothesis test
for the log rank test is:

$$
\begin{align*}
H_0 &: \text{in terms of survivability, there is no difference between the two groups} \\
H_1 &: \text{there is a survival difference between the two groups} 
\end{align*}
$$

```{r}
survdiff(Surv(time, status) ~ sex, data = lung)
```

Using an $\alpha = 0.05$ and the same data as the above Kaplan-Meier
curve comparing sex, the $H_0$ hypothesis is rejected as there is enough
evidence to suggest there is a survival difference between males and
females.

While the Kaplan-Meier curve and estimator are useful, there is an
assumption of non-informative censoring when using them. This assumption
means that there is no additional information on the survivial
likelihood for a subject after they have been censored. If that
assumption is violated, then Kaplan-Meier curves will not be estimating
the correct equation [@klein2005].

@@

To generate the CPH model, a survival object must first be created.
Using the R package `survival`, the survival formula can be generated by
running `Surv(time, event) ~ x`. That formula is then applied to the R
function `coxph(formula, data=)` which will generated the CPH model
[@R-base].

@@

For a CPH model to be valid, the following assumptions must be met
either from the raw data or through transformation.

@@

Evaluating a CPH model requires examining the predictors to determine
which ones are significant and generating the concordance index which
will determine how accurate the model is at predicting survival.

@@

TBD

@@

```{r}
#packages
library(tidyverse)
library(gridExtra)
library(survival)
library(survminer)
library(ggpubr)
library(magrittr)
library(ggfortify)
library(knitr)

@@

Example for one predictor

```{r}
martingale <- resid(cox, type = "martingale")
